---
title: "StormData"
author: "Matt Ralston"
date: "February 27, 2016"
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[LE,RO]{}
- \fancyfoot[LE,RO]{\thepage}
output: pdf_document
---

## Synopsis/Abstract

## Introduction
The basic goal of this assignment was to explore the NOAA Storm Database and answer some basic questions about severe weather events. Data were provided from the National Weather Service: [link](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2). 

## Questions
1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?


## Data Processing
```{r}
stormdata<-read.csv2("C:/Users/ralstonm/Projects/RepData_PeerAssessment2-master/repdata-data-StormData.csv.bz2",sep=",")
# Set the fatalities to be a numeric vector instead of a factor
stormdata$FATALITIES <- as.numeric(stormdata$FATALITIES)
# Select only the valid states
stormdata <- stormdata[stormdata$STATE %in% state.abb,]
# Replace events
stormdata$event <- NA
for (n in 1:nrows(stormdata)) {
  event <- tolower(stormdata[n,]$event)
  if (grepl('blizzard',event,ignore.case=T)) {
    event <- "Blizzard"
  } else if (grepl('cold',event,ignore.case=T)) {
    event <- "Cold"
  } else if (grepl('current',event,ignore.case=T)) {
    event <- "Sea"
  } else if (grepl('dry',event,ignore.case=T)) {
    event <- "Heat"
  } else if (grepl('fire',event,ignore.case=T)) {
    event <- "Fire"
  } else if (grepl('fld',event,ignore.case=T)) {
    event <- "Flood"
  } else if (grepl('flood',event,ignore.case=T)) {
    event <- "Flood"
  } else if (grepl('funnel',event,ignore.case=T)) {
    event <- "Tornado"
  } else if (grepl('freez',event,ignore.case=T)) {
    event <- "Cold"
  } else if (grepl('frost',event,ignore.case=T)) {
    event <- "Cold"
  } else if (grepl('hail',event,ignore.case=T)) {
    event <- "Hail"    
  } else if (grepl('heat',event,ignore.case=T)) {
    event <- "Heat"
  } else if (grepl('hot',event,ignore.case=T)) {
    event <- "Hot"
  } else if (grepl('hurricane',event,ignore.case=T)) {
    event <- "Hurricane"
  } else if (grepl('hypo',event,ignore.case=T)) {
    event <- "Cold"
  } else if (grepl('ice',event,ignore.case=T)) {
    event <- "Ice"
  } else if (grepl('light',event,ignore.case=T)) {
    event <- "Storm"
  } else if (grepl('precip',event,ignore.case=T)) {
    event <- "Storm"
  } else if (grepl('rain',event,ignore.case=T)) {
    event <- "Storm"
  } else if (grepl('shower',event,ignore.case=T)) {
    event <- "Storm"
  } else if (grepl('slide',event,ignore.case=T)) {
    event <- "Avalanche"
  } else if (grepl('snow',event,ignore.case=T)) {
    event <- "Snow"
  } else if (grepl('storm',event,ignore.case=T)) {
    event <- "Storm"
  } else if (grepl('summary',event,ignore.case=T)) {
    event <- "Unknown"
  } else if (grepl('surf',event,ignore.case=T)) {
    event <- "Surf"
  } else if (grepl('surg',event,ignore.case=T)) {
    event <- "Flood"
  } else if (grepl('swell',event,ignore.case=T)) {
    event <- "Surf"
  } else if (grepl('thun',event,ignore.case=T)) {
    event <- "Storm"
  } else if (grepl('tide',event,ignore.case=T)) {
    event <- "Flood"
  } else if (grepl('torn',event,ignore.case=T)) {
    event <- "Tornado"
  } else if (grepl('torrent',event,ignore.case=T)) {
    event <- "Precipiration"
  } else if (grepl('tropic',event,ignore.case=T)) {
    event <- "Tropical Storm"
  } else if (grepl('tstm',event,ignore.case=T)) {
    event <- "Wind"
  } else if (grepl('warm',event,ignore.case=T)) {
    event <- "Heat"
  } else if (grepl('waterspout',event,ignore.case=T)) {
    event <- "Tornado"
  } else if (grepl('wet',event,ignore.case=T)) {
    event <- "Precipitation"
  } else if (grepl('wind',event,ignore.case=T)) {
    event <- "Wind"
  } else if (grepl('winter',event,ignore.case=T)) {
    event <- "Cold"
  } else if (grepl('volcan',event,ignore.case=T)) {
    event <- "Volcano"
  } else if (event == "?") {
    event <- "Unknown"
  } else {
    event <- gsub('^ ', gsub(' $',event,rep=""), rep="")
  }
  stormdata[n,]$event <- event
}


exp <- c(100,1000,1000000,1000000000,1,2,3,4,5,6,7,8,9)
names(exp) <- c("H","K","M","B","1","2","3","4","5","6","7","8","9")
exponent <- function(x) {
  if (toupper(x) %in% names(exp)) {
    x <- exp[toupper(x)]
  } else {
    x <- 1
  }
  return(x)
}

for (e in names(exp)) {
  propReplaced <- stormdata$PROPDMGEXP == e
  cropReplaced <- stormdata$CROPDMGEXP == e
  stormdata[propReplaced,]$propdmg <- as.numeric(stormdata$PROPDMG[propReplaced])*exponent(e)
  stormdata[cropReplaced,]$cropdmg <- as.numeric(stormdata$CROPDMG[cropReplaced])*exp[e]
}


```

## Results
1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
```{r}
fatalitiesByEvent <- aggregate(stormdata$FATALITIES,by=list(event=stormdata$event),sum)
names(fatalitiesByEvent)<-c("Event","Fatalities")
fatalitiesByEvent <- fatalitiesByEvent[order(fatalitiesByEvent$Fatalities,decreasing = T),]
# Print the top 10 most fatal event types.
head(fatalitiesByEvent,n=10)
ggplot(fatalitiesByEvent)+geom_bar(aes(x=Event,y=Fatalities))
```

2. Across the United States, which types of events have the greatest economic consequences?

```{r}
cropByEvent <- aggregate(stormdata$cropdmg,by=list(stormdata$event),sum)
propByEvent <- aggregate(stormdata$propdmg,by=list(stormdata$event),sum)
names(cropByEvent)<-c("Event","CropDamage")
names(propByEvent)<-c("Event","PropertyDamage")
cropByEvent <- cropByEvent[order(cropByEvent$CropDamage,decreasing=T),]
propByEvent <- propByEvent[order(propByEvent$PropertyDamage,decreasing=T),]
print("Top 10 Crop Damaging Events:")
head(cropByEvent, n=10)
print("Top 10 Property Damaging Events:")
head(propByEvent, n=10)
```
